### 2023-05-18 docs/minutes/slides_19_05/document.tex
slides

@@

\begin{frame}
	Full NS search for \alert{all parameters} \newline 
	$\bar{\theta} = [\bar{\theta}_{\rm GW}, \, f, \, (t=0) \, \dot{f} \, (t=0) ]$ \newline 
	with $h = 10^{-10}$, $\sigma_m = 10^{-11}$ \newline 
	Note: pulsar distance is no longer part of model. $\gamma$ doesn't matter

@@

\begin{frame}
		\includegraphics[width=0.9\textwidth,height=0.9\textwidth]{images/canonical_mixed_result}
\end{frame}

\begin{frame}
	\includegraphics[width=0.9\textwidth,height=0.9\textwidth]{images/canonical_mixed_f_result}
\end{frame}

\begin{frame}
	Next steps:
	\begin{itemize}
		\item Bayes factors vs. strain (currently running)
		\item Process noise $\sigma_p$
		\item Looser priors on $f$, $\dot{f}$? Do we need to?
		\item Can we add PSR terms back in? Do we need to?
	\end{itemize}
\end{frame}

@@

		AM: What changed such that now everything works?

		\begin{itemize}
			\item Dropping PSR terms to smooth likelihoods
			\item \textit{Increasing} measurement noise  
			\item Don't need too many live points
			\item Non-default Bilby sampler settings 
			\begin{itemize}
				\item e.g. docs recommend `sample=act-walk`or `sample=acceptance-walk`
				\item More success with `sample=rwalk-dynesty` (not `sample=rwalk` which is the Bilby implementation)
			\end{itemize}
		\item Be mindful of float errors - Python/Numpy/Bilby Nans get propagated, rather than error e.g. Fortran.
		\end{itemize}

@@

@@

		\begin{itemize}
		\item \alert{Problem} = overly narrow,biased posteriors
		\item Cause: noisy, likelihoods which are "locally Gaussian". Sampling gets stuck in a local optima and infers posteriors widths based on that optima. 
%		\item Require $n_{live} > \frac{\rm prior \, volume}{\rm mode \,  footprint} = \frac{X}{\sigma^2}$
%		\begin{itemize}
%			\item For $\delta$ require $n_{live} > \frac{2 \pi}{10^{-5}} \sim 10^5 $
%			\item For $\omega$ require $n_{live} > \frac{10^4}{10^{-11}} \sim 10^7 $ 
%			\item Currently $n_{live} = 500$ parallelized on 32 cores $\sim 700$ minutes
%			\item Runtime scales $\mathcal{O}(n_{live}) \rightarrow 8000 $ hours for required $\delta$ sampling (just under 1 year!)
%
%		\end{itemize}
	\end{itemize}

@@

\begin{frame}{}
	Potential solutions:

@@

\begin{frame}{Option 2: Drop PSR terms from model?}

	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/canonical_mixed_result}

	OK, but some biases. c.f. location of the maxima on the previous slide
\end{frame}

	Lets proceed using this `Earth terms' model for now.

   So far we have been considering inference of only the GW terms $\bar{\theta}_{\rm GW}$. It is important to be able to infer the pulsar terms $\bar{\theta}_{\rm PSR}$.

   Lets try and include $f \, (t=0)$ and $\dot{f} \, (t=0)$ in our nested sampling.

   We have a uniform prior of $\pm 1 \%$ about the true value for these two parameters

\begin{frame}{Earth terms model $\bar{\theta}_{\rm GW}$ and $f \, (t=0)$, $\dot{f} \, (t=0)$}
	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/canonical_mixed_result}
\end{frame}

>>>>

\definecolor{links}{HTML}{2A1B81}

\hypersetup{colorlinks,linkcolor=,urlcolor=links}

@@

\begin{frame}
	Full NS search for \alert{all parameters} \newline 
	$\bar{\theta} = [\bar{\theta}_{\rm GW}, \, f_0, \, \dot{f}_0]$ \newline 
	with $h = [10^{-10}, 10^{-12}]$, $\sigma_m = 10^{-11}$ \newline 
	Note: pulsar distance is no longer part of model. $\gamma$ doesn't matter

@@

\begin{frame}
		\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical_mixed_result_10}
		\only<1>{%
			\tikz[overlay,remember picture]
			\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-10}$,  $\bar{\theta}_{\rm GW}$};

		}
			\only<1>{%
		\tikz[overlay,remember picture]
		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=2.5cm]current page.center){Note some systematics in $\alpha$, $\psi$};

	}
\end{frame}

\begin{frame}
	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_result}
	\only<1>{%
		\tikz[overlay,remember picture]
		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$,$\bar{\theta}_{\rm GW}$};

	}
	\only<1>{%
		\tikz[overlay,remember picture]
		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=2.5cm]current page.center){Note $h -\iota$ trade-off};

	}

\begin{frame}
\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_f_result}
\only<1>{%
	\tikz[overlay,remember picture]
	\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$, $f_0$, all (47) PSRs};

}

\begin{frame}
	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_fdot_result}
	\only<1>{%
		\tikz[overlay,remember picture]
		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$, $\dot{f}_0$, all (47) PSRs};

	}
\end{frame}

\begin{frame}
	Next steps:
	\begin{itemize}
		\item Bayes factors vs. strain (currently running)
		\item Process noise $\sigma_p$ (next slides)
		\item $[h, \iota] \rightarrow [h_{+}, h_{\times}]$  (discussed, easy, helpful?)
		\item Can we add PSR terms back in? Do we need to?
		\item Robustness for different $\bar{\theta}_{\rm GW}$ (overkill? canonical example sufficient?)
	\end{itemize}
\end{frame}

\begin{frame}{Setting the process noise}

Timing noise model of \href{https://arxiv.org/abs/1010.4794}{Shannon \& Cordes, 2010}
\begin{eqnarray}
 \ln \sigma_p^{\rm TOA} [\mu s] = \ln \alert{C} + \alert{\alpha} \ln \nu [s^{-1}] + \alert{\beta} \ln \dot{\nu} [10^{-15} s^{-2}] + \alert{\gamma} \ln T [years]
\end{eqnarray}

For MSP, $\ln C \sim -20$, $\alpha \sim 1$, $\beta\sim2$, $\gamma \sim 2.4$

Taking $T = 1$ week, for our synthetic NANOGrav pulsars this puts:	 $$\sigma_p^{\rm TOA}  [s] \sim (10^{-19}, 10^{-17}, 10^{-13}) \, \, , \text{(min/median/max)}$$

Relate to frequency:
$$ \sigma_p^{\rm f} [Hz] = f \, \frac{\sigma_p^{\rm TOA}}{T}  \sim (10^{-23}, 10^{-21}, 10^{-16}) \, \, , \text{(min/median/max)}$$

Could hit float epsilon issues here... may need to scale state evolution equations, similar to heterodyning the measured data...

\begin{frame}{Rescaling to avoid float issues}
	State evolution equation (Ornstein–Uhlenbeck)
	\begin{eqnarray}
		\frac{df}{dt}= -\gamma [f - f_{\rm EM}(t)] + \dot{f}_{\rm EM} + \xi(t; \sigma_p)
	\end{eqnarray}
i.e. 
		\begin{eqnarray}
		df = a(f,t) dt + \sigma_p dW
	\end{eqnarray}

Can solve numerically via e.g. Euler-Maruyama

Partition interval $[0,T]$ into $N$ intervals, each width $\Delta t = T/N$ 
\begin{eqnarray}
	f_{n+1} = f_{n} + a(f_n, \tau_n) \Delta t + \sigma_p \Delta W_n
\end{eqnarray}
\begin{eqnarray}
		\mathcal{O}(10^2) = \mathcal{O}(10^2) + \mathcal{O}(10^{-15}) \cdot \Delta t  + \mathcal{O}(10^{-20}) \Delta W_n
\end{eqnarray}

Practically: synthetic data (float64) generated with e.g. $\sigma_p = 10^{-19}$ = data with e.g. $\sigma_p = 10^{-20}$

\begin{frame}{}
	\includegraphics[width=0.90\textwidth,height=0.60\textwidth]{images/sigma_p_non_hetero}

\begin{frame}{}
	Let $f \, ' = f - f_{\rm EM}$
	\begin{align}
		\frac{df \, '}{dt}&= \frac{df}{dt} - \frac{df_{\rm EM}}{dt} \\
		                    &= \frac{df}{dt} - \frac{df_{\rm EM}}{dt} \\
		                    &= - \gamma f \, '  + \xi 
	\end{align}
	Now numerical solution looks like
	\begin{eqnarray}
				\mathcal{O}(10^{-12}) = \mathcal{O}(10^{-13}) \cdot \mathcal{O}(10^{-12}) \cdot \Delta t  + \mathcal{O}(10^{-20})
	\end{eqnarray}
	\begin{eqnarray}
	\mathcal{O}(1) = \mathcal{O}(10^{-13}) \cdot \Delta t  + \mathcal{O}(10^{-8})
\end{eqnarray}

	Practically: synthetic data (float64) generated with e.g. $\sigma_p = 10^{-19}$ $\neq$ data with e.g. $\sigma_p = 10^{-20}$

\begin{frame}{}
	\includegraphics[width=0.90\textwidth,height=0.60\textwidth]{images/sigma_p_hetero}

\begin{frame}{Heterodyne state and measurement?}

	\begin{itemize}
		\item Heterodyne the state: $f \, ' = f - f_{\rm EM}$ 
		\item Heterodyne the measurement: $f \, '_{\rm M}  = f_{\rm M} - f^{\, *}_{\rm EM}$ 
	\end{itemize}

	$f_{\rm EM}$ is the exact, true solution parametrised by the unknowns $f_0, \dot{f}_0$

	 $f^{\, *}_{\rm EM}$ is a guess based on the pulsar ephemeris.

	 For our case where we have synthetic data we can set $f_{\rm EM} = f^{\, *}_{\rm EM}$, but obviously not true generally - we are trying to estimate $f_{\rm EM}$!

	We can then write the measurement equation as
	\begin{eqnarray}
		f_{\rm M} = (1 - X)f + N_{\rm G} 
	\end{eqnarray}
\begin{eqnarray}
	f \, '_{\rm M} = (1 - X)f \, ' - X f_{\rm EM} + \alert{(f_{\rm EM} - f^{\, *}_{\rm EM})} +  N_{\rm G}
\end{eqnarray}

	\begin{itemize}
		\item Linear \checkmark (control vector has shifted from state equation to measurement equation)
		\item Scales: $\mathcal{O}(h) = \mathcal{O}(10^{-12}) - \mathcal{O}(h)\cdot\mathcal{O}(10^2) + \alert{\mathcal{O}(0)} + \mathcal{O}(10^{-11})$ \checkmark
	\end{itemize}

Insert corner plot here

@@

		AM: \textit{"What changed such that now everything works?"}

		\begin{itemize}
			\item Dropping PSR terms to smooth likelihoods
			\item \textit{Increasing} measurement noise  
			\item Don't need too many live points
			\begin{itemize}
				\item $n_{live} > $ number of parameters 
				\item $n_{live} \times \frac{\rm mode \, volume}{\rm prior \, volume} > 1 $ i.e. at least 1 live point per mode
				\item Runtime scales $\mathcal{O}(n_{live})$
				\item Posterior/Evidence uncertainties scale as $\mathcal{O}(1 / \sqrt{n_{live}})$
			\end{itemize}

			\item Non-default Bilby sampler settings 
			\begin{itemize}
				\item e.g. docs recommend `sample=act-walk`or `sample=acceptance-walk`
				\item More success with `sample=rwalk-dynesty` (not `sample=rwalk` which is the Bilby implementation)
				\item bound = multi(dynesty), not the Bilby version 
			\end{itemize}
		\item Be mindful of float errors - Python/Numpy/Bilby NaNs get propagated, rather than error e.g. Fortran.
		\end{itemize}

@@

\begin{frame}{More likelihood curves}

@@

		\begin{itemize}
		\item \alert{Problem} = overly narrow,biased posteriors
		\item Cause: noisy, likelihoods which are "locally Gaussian". Sampling gets stuck in a local optima and infers posteriors widths based on that optima.

%		\end{itemize}
	\end{itemize}

@@

\begin{frame}{Potential solutions to noisy likelihoods}

@@
